Estimating the distribution of block sizes as a function of average transaction rates

### 2015-10-22 paper/paper.tex
Fixed some typos and improved some explanation

\section{Introduction}
The optimal maximum size of Bitcoin blocks is an area of active controversy. When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks\citep{eyal2014majority}.%\footnote{http://arxiv.org/abs/1311.0243}
 When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin. The issue may be with us for some time; even assuming that current disagreements over raising the 1MB cap are resolved, if the Bitcoin network keeps growing and as computing power increases, the issue will need to be re-litigated from time to time. One proposal, BIP 100, would subject the decision to a continuous voting process by miners indefinitely into the future.

@@

\section{Verifying the variability of transaction rates}
Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\delta_k \equiv \tau_{k} - \tau_{k-1} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$. The longer $\delta_k$ the more time for transactions to add up before being added to the block chain as part of block $k$.

@@

A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale. In order to test this assumption we fit a gamma distribution to the block sizes of various pieces of the block chain. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$. Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant. We focus on the most recent 5000 blocks in our database ending with block 303151 --- we are in the process of downloading the entire chain.  For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$. This prior on $\alpha$ implies that the prior median is $\alpha=1$ so that the null hypothesis is favored relatively strongly by the prior. We fit the model to all 5000 blocks, the first 1000 blocks, the second 1000 blocks, etc., and the first 500 blocks, the second 500 blocks, etc. The results are in Table \ref{tab:gammafit}.

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how how the amount of transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider non-negative infinite activity L'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.

The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$ and for which $x_\tau$ is inverse Gaussian distributed for all $\tau>0$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_{k-1} + \delta_k}$ without resorting to approximation methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar model that avoids numerical problems is the lognormal distribution, which we will use.

@@

Given an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways. The simplest way is to compute parameter estimates based on the posterior, e.g. posterior means or posterior medians, then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant. In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern. This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{beta}$ where $\hat{\beta}$ is an estimate of $\beta$.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution \citep{gelman2014bayesian}. In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct. Explicitly constructing this prior would be involved and maybe intractable or even impossible. Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.

@@

Under a fixed mean transaction rate, the probability of a full block increases as as $\beta$ decreases, especially for smaller mean transaction rates. For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$. This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes. When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect. What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills up no matter the expected value of $\gamma$.

These estimates seem rather high considering no block have filled up yet. Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1. Even at a constant transaction rate of 0.02, the probability of a full block is 0.067. In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$. Since no block has filled up yet, clearly something is happening that prevents blocks from filling up. One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate. Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up. Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average. Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.

>>>>

\section{Introduction}
The optimal maximum size of Bitcoin blocks is an area of active controversy. When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks \citep{eyal2014majority}.%\footnote{http://arxiv.org/abs/1311.0243}
 When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin. The issue may be with us for some time; even assuming that current disagreements over raising the 1MB cap are resolved, if the Bitcoin network keeps growing and as computing power increases, the issue will need to be re-litigated from time to time. One proposal, BIP 100, would subject the decision to a continuous voting process by miners indefinitely into the future.

@@

\section{Verifying the variability of transaction rates}
Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for an increment $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed, i.e. $\delta_k \equiv \tau_{k} - \tau_{k-1} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$ where $\tau_k$ is the arrival time of the $k$'th block. We use the rate parameterization of the exponential distribution so that if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$. The longer $\delta_k$ the more time for transactions to add up before being added to the block chain as part of block $k$.

@@

A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale. In order to test this assumption we fit a gamma distribution to the block sizes of various pieces of the block chain. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$, $\alpha>0$, and $\beta>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$. The gamma function in the denominator is defined as $\Gamma(x) = \int_0^\infty x^{t-1}e^{-x}dx$. Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant. We focus on the most recent 5000 blocks in our database ending with block 303151 --- we are in the process of downloading the entire chain.  For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$. This prior on $\alpha$ implies that the prior median is $\alpha=1$ so that the null hypothesis is favored relatively strongly by the prior. We fit the model to all 5000 blocks, the first 1000 blocks, the second 1000 blocks, etc., and the first 500 blocks, the second 500 blocks, etc. The results are in Table \ref{tab:gammafit}.

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how the amount of transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider nonnegative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. The gamma process is called an infinite activitive process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.

The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. More generally the gamma distribution is infinitely divisible so that if $x\sim G(\alpha,\beta)$ then for any integer $n>0$ there exist iid random variables $y_1,y_2,\dots,y_n$ such that $y_1+y_2+\dots+y_n$ and $x$ have the same distribution. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process for which the distribution of $x_\tau$ is in the same class as the distribution for $x_1$ for all $\tau>0$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar distribution that avoids numerical problems is the lognormal, which we will use.

@@

Given an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways. The simplest way is to compute parameter estimates based on the posterior, e.g. posterior means or posterior medians, then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant. In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern. This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{\beta}$ where $\hat{\beta}$ is an estimate of $\beta$.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. In particular, 2) implies that the variance of transaction rates is always proportional to the mean, i.e. it is $\gamma/\beta$. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution \citep{gelman2014bayesian}. In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. In other words, it still assumes that the variance of transaction rates is still $\gamma/\beta$, but now $\gamma$ and $\beta$ are also stochastic with $\beta$ distributed according to its marginal posterior distribution and $\gamma|\beta$ distributed according to a shifted version of its conditional posterior given $\beta$.

 Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct. Explicitly constructing this prior would be involved and maybe intractable or even impossible. Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.

@@

Under a fixed mean transaction rate, the probability of a full block increases as $\beta$ decreases, especially for smaller mean transaction rates. For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$. This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes. When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect. What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills given a particular value for the expected value of $\gamma$.

These estimates may seem rather high considering blocks only recently began filling up. Recent transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1. With a constant transaction rate of 0.02, the probability of a full block is 0.067. This seems small, but in 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$. The apparent paradox is solved by acknowledging a soft limit that Bitcoin miners impose on block sizes. While the Bitcoin protocol limits block sizes to 1 MB, Bitcoin miners have the ability to set their own block size limit lower. Miners have an incentive to keep blocks small since that reduces the chance of an orphan block --- a mined block which is not accepted by the blockchain --- so they tend to exercise this option. Further, the default setting is to limit block sizes to 750 KB.
